Standard Candle Cosmology With Incomplete Spectroscopic Classification

### 2016-02-23 doc/toy.tex
rewrite of toy note and cleaning up of some hanging changes

The ambition is to construct a model to determine the likelihood of DES-like
supernova data.
The conceptual structure of the model is shown in Figure~\ref{pgm:fig}. 
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=6.5in]{../results//nodes_pgm.pdf} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.  
   \label{pgm:fig}}
\end{figure}

In the toy model there are three types of data that
may be observed for each supernova:
\begin{itemize}
\item $c_o$: Counts at peak brightness.
\item $z_o$: Redshift.
\item $T_o$: Transient type.
\end{itemize}

There are selection criteria $S_c$ and $S_T$ for the inclusion of an object in the
sample and for a successul typing.
In what follows we assume that the typed objects are drawn from the same underlying
distribution as the sample $S_c=S_T$, and that the selection criterion is a brightness
threshold $S_c = c_o \ge c_T$.  (The spectroscopic sample $S_T$ 
eventually should depend on the
photometry and the quality of the spectroscopic data.)

For this preliminary discussion the top-level parameters are denoted by $X$,
latent parameters $L$ and $T$.

@@

\subsection{Case Typed}
For Case Typed the likelihood is
\begin{align}
\mathcal{L}(T_o,z_o,c_o | S_c, S_T, z, X) & =  \int dL \sum_i p(T_o,z_o,c_o | T_i, L, S_c, S_T, z, X) p(L |  T_i,  S_c, S_T, z, X) P(T_i|S_c, S_T, z, X).
\end{align}
We assume that the observations give a perfect measurement of type and redshift,
so that
\begin{align}
p(T_o|T) & =\delta(T_o-T)\\
p(z_o|z) & =\delta(z_o-z);
\end{align}
practically $T$ and $z$ are no longer probabilistic but are fixed.
Then
\begin{align}
\mathcal{L}(T_o,z_o,c_o | S_c, S_T, z, X) & =  \int dL\, p(c_o | T=T_o, L, S_c, S_T, z=z_o, X) p(L| T=T_o, S_c, S_T, z=z_o, X)  P(T_o|S_c, S_T, z=z_o, X) \\
&= \int dL\, \frac{ p(c_o, S_c, S_T | T=T_o, L, z=z_o, X) }{P(S_c, S_T, | T=T_o, L,  z=z_o, X) }
\frac{p(L, S_c, S_T | T=T_o, z=z_o, X)}{P(S_c, S_T| T=T_o,  z=z_o, X)}
\frac{p(T_o, S_c, S_T | z=z_o, X)}{P(S_c, S_T| z=z_o, X)} \\
&= \frac{\int dL\, p(L|T_o, z=z_o, X) P(T_o|z=z_o, X) p(c_o, S_c, S_T | T=T_o, L, z=z_o, X)}{P(S_c, S_T| z=z_o, X)}\\
&= \frac{\int dL\, p(L|T_o, z=z_o, X) P(T_o|z=z_o, X) p(c_o | T=T_o, L, z=z_o, X)}{P(S_c, S_T| z=z_o, X)}.
\end{align}

@@

In the case of a single count measurement per transient and neglecting the typing selection criteria,
\begin{align}
 p(c_o, S_c, S_T | T=T_o, L, z=z_o, X) &=  p(c_o, c_o \ge  c_T | T=T_o, L, z=z_o, X) \\
 &= \begin{cases}
   p(c_o | T=T_o, L, z=z_o, X) & \text{if } c_o \ge c_T \\
   0 & \text{if }  c_o < c_T
 \end{cases}
\end{align}
so that the integral in brackets is
\begin{equation}
\int dc_o \, p(c_o,S_c, S_T | T=T_i, L, z=z_o, X) = \int_{c_T}^\infty dc_o\,  p(c_o | T=T_i, L, z=z_o, X).
\end{equation}

For the more general case, suppose a transient has $N$ independent count measurements\footnote{The independence of count measurements  depends on the flux extraction method; regardless the
noise of a single observation is expected to be dominated by its own photon noise making this
supposition a good approximation.} and $k$ of those must be above threshold in order to be
included in the sample.
Consider arrays of length $N$ in which each element has a value of True or False: there are $2^N$ possible configurations for this array
of which $\sum_{i=k}^N {N \choose i}$ have $k$ or more True's.  Let the set $\mathcal{G}$ represent the latter possible configurations.
Then
\begin{align}
p(c_o, S_c| T=T_o, L, z=z_o, X) &= \sum_{g \in \mathcal{G}} \prod_{\alpha} p(c_{o, \alpha}, g | T=T_o, L, z=z_o, X)\\
 &=  \sum_{g \in \mathcal{G}}  \prod_{\alpha} \begin{cases}
   p(c_{o,\alpha} | T=T_o, L, z=z_o, X) & \text{if } c_{o,\alpha} \ge c_T == g_\alpha \\
   0 & \text{if }  c_{o,\alpha} \ge c_T\  != g_\alpha
 \end{cases}
\end{align}
and the normalization integral is composed of the sum over all possible configurations of the integrated probability of the truncated distributions
of the measured counts.  This simplifies to the the single-count case considered earlier in the case of $N=1$, $k=1$.

@@

For the redshift measurement, we take a set of potential host galaxies, where galaxy $j$
are redshift $z_j$ has probability $p_j$ of being the host
\begin{equation}
p(z_o|S_c, S_T, z, X) = \sum_j p_j\delta(z_j-z).
\end{equation}
Then
\begin{align}
\mathcal{L}(z_o,c_o | S_c, z, X) & =  \sum_j p_j \int dL \,\sum_{i} p(c_o | T_i, L, S_c, z=z_j, X) p(L |  T_i,  S_c,  z=z_j, X) P(T_i|S_c, z=z_j, X) \\
&=  \frac{\sum_j p_j \int dL \, p(c_o, S_c | L, z=z_j, X) \sum_{i}  p(L|T_i, z=z_j, X) P(T_i|z=z_j, X)   }{P(S_c| z=z_j, X)}\\
&=  \frac{\sum_j p_j \int dL \, p(c_o | L, z=z_j, X) \sum_{i}  p(L|T_i, z=z_j, X) P(T_i|z=z_j, X)   }{P(S_c| z=z_j, X)},
\end{align}
where we use the fact that the counts do not directly depend on type.

@@

@@

\subsection{Cosmology}
The Flat $w$CDM cosmology.

\begin{align}
\Omega_M & \sim \ln{\mathcal{N}}(\ln{0.28},0.1)\\
w_0 & \sim \mathcal{N}(-1,0.05)
\end{align}

\subsection{Calibration}
Photometric zeropoints for each band.
\begin{equation}
Z \sim \mathcal{N}(0,0.02)
\end{equation}

@@

\subsection{Sample Selection Normalization}
Anticipating the use of available statistical packages, the likelihoods have been constructed
from standard distribution functions.  The truncated distributions from sample selection
are accounted by the normalization 
terms $P(S_c, S_T| z=z_o, X)$, $P(S_c| z=z_o, X)$.  Although they do not appear
in the PGM, they are treated as nodes who contribute 
$-\ln{P(S_c, S_T| z=z_o, X)}$ to the log-likelihood, where the negative sign reflects that this is a normalization
factor.

\subsection{Implementation Notes}
The following integral appears in the normalization
\begin{equation}
\int dL p(L|T_i, z=z_o, X)  \left[\int_{c_T}^{\infty} dc_o  p(c_o | T=T_i, L, z=z_o, X)\right].
\end{equation}
We actually work with $\ln{L_X}$ and $\sigma_X$.  Setting
\begin{equation}
n=\frac{\ln{L}-\ln{L_X}}{\sigma_X}
\end{equation}
the integral becomes
 \begin{equation}
\sigma_X \int dn p(n\sigma_X + \ln{L_X} |T_i, z=z_o, X)  \left[\int_{c_T}^{\infty} dc_o  p(c_o | T=T_i, L_Xe^{n\sigma_X}, z=z_o, X)\right].
\end{equation}
The two probability distributions are normal so that this reduces to
 \begin{equation}
\frac{1}{2\sqrt{2\pi}} \int dn\, e^{-0.5n^2} \left(1-erf\left(\frac{(c_T - c_\star)}{\sqrt{2}\sigma_c}\right)\right),
\end{equation}
where
\begin{equation}
c_\star = \frac{e^{\ln{L_X}+n\sigma_X}}{4\pi d_L^2}10^{Z/2.5}.
\end{equation}
\end{document}

>>>>

The ambition is to construct a model to determine the likelihood of DES-like
supernova data.  We start with a toy model and plan to evolve its complexity to
have the fidelity necessary for the DES analysis.
The conceptual structure of the model is shown in Figure~\ref{pgm:fig}. 
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=6.5in]{../results//nodes_pgm.pdf} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.  
   \label{pgm:fig}}
\end{figure}

In the toy model there are three types of data that
may be observed for each supernova:
\begin{itemize}
\item $c_o$: Counts at peak brightness, one or more independent measurements.
\item $z_o$: Redshift.
\item $T_o$: Transient type.
\end{itemize}

There are selection criteria $S_c$ and $S_T$ for the inclusion of an object in the
sample and for a successul typing.
In what follows we assume that the typed objects are drawn from the same underlying
distribution as the sample $S_c=S_T$, and that the selection criterion depends on a  brightness
threshold $S_c = c_o \ge c_T$.  (The spectroscopic sample $S_T$ 
eventually should depend on the
photometry and the quality of the spectroscopic data.)

For this preliminary discussion the top-level parameters are denoted by $X$,
the latent parameters are $L$ and $T$ (for luminosity and type).

@@

\subsection{Case Typed}
For Case Typed the likelihood is
\begin{align}
\mathcal{L}(T_o,z_o,c_o | S_c, S_T, z, X) & =  \int dL \sum_i p(T_o,z_o,c_o | T_i, L, S_c, S_T, z, X) p(L |  T_i,  S_c, S_T, z, X) P(T_i|S_c, S_T, z, X).
\end{align}
We assume that the observations give a perfect measurement of type and redshift,
so that
\begin{align}
p(T_o|T) & =\delta(T_o-T)\\
p(z_o|z) & =\delta(z_o-z);
\end{align}
$T$ and $z$ are no longer probabilistic but are fixed.
Then
\begin{align}
\mathcal{L}(T_o,z_o,c_o | S_c, S_T, z, X) & =  \int dL\, p(c_o | T=T_o, L, S_c, S_T, z=z_o, X) p(L| T=T_o, S_c, S_T, z=z_o, X)  P(T_o|S_c, S_T, z=z_o, X) \\
&= \int dL\, \frac{ p(c_o, S_c, S_T | T=T_o, L, z=z_o, X) }{P(S_c, S_T, | T=T_o, L,  z=z_o, X) }
\frac{p(L, S_c, S_T | T=T_o, z=z_o, X)}{P(S_c, S_T| T=T_o,  z=z_o, X)}
\frac{p(T_o, S_c, S_T | z=z_o, X)}{P(S_c, S_T| z=z_o, X)} \\
&= \frac{\int dL\, p(L|T_o, z=z_o, X) P(T_o|z=z_o, X) p(c_o, S_c, S_T | T=T_o, L, z=z_o, X)}{P(S_c, S_T| z=z_o, X)}\\
&= \frac{\int dL\, p(L|T_o, z=z_o, X) P(T_o|z=z_o, X) p(c_o | T=T_o, L, z=z_o, X)}{P(S_c, S_T| z=z_o, X)}.
\end{align}

@@

@@

For the redshift measurement, we take a set of potential host galaxies, where galaxy $j$
are redshift $z_j$ has probability $p_j$ of being the host
\begin{equation}
p(z_o|S_c, S_T, z, X) = \sum_j   p_j\delta(z_j-z).
\end{equation}
Then
\begin{align}
\mathcal{L}(z_o,c_o | S_c, S_T, z, X) & =  \sum_j p_j \int dL \,\sum_{i} p(c_o | T_i, L, S_c, S_T, z=z_j, X) p(L |  T_i,  S_c, S_T,  z=z_j, X) P(T_i|S_c, S_T, z=z_j, X) \\
&=  \frac{\sum_j p_j \int dL \, p(c_o, S_c, S_T | L, z=z_j, X) \sum_{i}  p(L|T_i, z=z_j, X) P(T_i|z=z_j, X)   }{P(S_c, S_T| z=z_j, X)}\\
&=  \frac{\sum_j p_j \int dL \, p(c_o | L, z=z_j, X) \sum_{i}  p(L|T_i, z=z_j, X) P(T_i|z=z_j, X)   }{P(S_c, S_T| z=z_j, X)},
\end{align}
where we use the fact that the counts do not directly depend on type.

@@

\subsection{Sample Selection}
Both the Typed and Not Typed cases have in their normalizations the term
\begin{equation}
\int dc_o \, p(c_o,S_c, S_T | T=T_i, L, z=z_o, X),
\end{equation}
which is the fraction of objects whose counts satisfy the sample selection criteria.

In the case of a single count measurement per transient and neglecting the typing selection criteria,
\begin{align}
 p(c_o, S_c, S_T | T=T_o, L, z=z_o, X) &=  p(c_o, c_o \ge  c_T | T=T_o, L, z=z_o, X) \\
 &= \begin{cases}
   p(c_o | T=T_o, L, z=z_o, X) & \text{if } c_o \ge c_T \\
   0 & \text{if }  c_o < c_T
 \end{cases}
\end{align}
so that the integral is
\begin{equation}
\int dc_o \, p(c_o,S_c, S_T | T=T_i, L, z=z_o, X) = \int_{c_T}^\infty dc_o\,  p(c_o | T=T_i, L, z=z_o, X).
\end{equation}
Our count measurements come from a Normal distribution, so that integral is simply related to the error function.

For the more general case, suppose a transient has $N$ independent count measurements\footnote{The independence of count measurements  depends on the flux extraction method; regardless the
noise of a single observation is expected to be dominated by its own photon noise making this
supposition a good approximation.} and $k$ of those must be above threshold in order to be
included in the sample.
Consider arrays of length $N$ in which each element has a value of True or False: there are $2^N$ possible configurations for this array
of which $\sum_{i=k}^N {N \choose i}$ have $k$ or more True's.  Let the set $\mathcal{G}$ represent the latter possible configurations.
Then
\begin{align}
p(c_o, S_c| T=T_o, L, z=z_o, X) &= \sum_{g \in \mathcal{G}} \prod_{\alpha} p(c_{o, \alpha}, g | T=T_o, L, z=z_o, X)\\
 &=  \sum_{g \in \mathcal{G}}  \prod_{\alpha} \begin{cases}
   p(c_{o,\alpha} | T=T_o, L, z=z_o, X) & \text{if } (c_{o,\alpha} \ge c_T) = g_\alpha\\
   0 & \text{if }  (c_{o,\alpha} \ge c_T)  \ne g_\alpha
 \end{cases}
\end{align}
and the normalization integral is composed of the sum over all possible configurations of the integrated probability of the truncated distributions
of the measured counts.  This simplifies to the the single-count case considered earlier in the case of $N=1$, $k=1$.
If the count measurement uncertainties are independent, which may not be exactly but is approximately the case when we have high
signal-to-noise references, the integral because a sum of products of normalized error functions.

@@

\subsection{Cosmology}
The Flat $w$CDM cosmological parameters are drawn from top-hat distributions
\begin{align}
\Omega_M & \sim  {\mathcal{U}}(0,1)\\
w_0 & \sim \mathcal{U}(-1.5, -0.1).
\end{align}

\subsection{Calibration}
Photometric zeropoints for each band.
\begin{equation}
Z \sim \mathcal{N}(0,\frac{\ln{10}}{2.5}0.01)
\end{equation}

@@

\subsection{Implementation Notes}
Anticipating the use of available statistical packages, the likelihoods have been constructed
from standard distribution functions.  The truncated distributions from sample selection
are accounted by the normalization 
terms $P(S_c, S_T| z=z_o, X)$, $P(S_c| z=z_o, X)$.  Although they do not appear
in the PGM, they are treated as nodes who contribute 
$-\ln{P(S_c, S_T| z=z_o, X)}$ to the log-likelihood, where the negative sign reflects that this is a normalization
factor.

The following integral appears in the normalization
\begin{equation}
\int dL p(L|T_i, z=z_o, X)  \left[\int_{c_T}^{\infty} dc_o  p(c_o | T=T_i, L, z=z_o, X)\right].
\end{equation}
We actually work with $\ln{L_X}$ and $\sigma_X$.  Setting
\begin{equation}
n=\frac{\ln{L}-\ln{L_X}}{\sigma_X}
\end{equation}
the integral becomes
 \begin{equation}
\sigma_X \int dn p(n\sigma_X + \ln{L_X} |T_i, z=z_o, X)  \left[\int_{c_T}^{\infty} dc_o  p(c_o | T=T_i, L_Xe^{n\sigma_X}, z=z_o, X)\right].
\end{equation}
The two probability distributions are normal so that this reduces to
 \begin{equation}
\frac{1}{2\sqrt{2\pi}} \int dn\, e^{-0.5n^2} \left(1-erf\left(\frac{(c_T - c_\star)}{\sqrt{2}\sigma_c}\right)\right),
\end{equation}
where
\begin{equation}
c_\star = \frac{e^{\ln{L_X}+n\sigma_X}}{4\pi d_L^2}10^{Z/2.5}.
\end{equation}

\end{document}
